---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
[![Travis-CI Build Status](https://travis-ci.org/bbuchsbaum/neurosurf.svg?branch=master)](https://travis-ci.org/bbuchsbaum/neurosurf)

# neurosurf

Under development

## Installation

You can install neurosurf from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("bbuchsbaum/neurosurf")
```

## Vignettes

See examples of use of `neurosurf` in the [vignettes](https://bbuchsbaum.github.io/neurosurf/articles/index.html).

## Curvature shading

`surfwidget()` includes an optional `curvature` argument supplying per vertex
curvature values. When omitted and the input is a `SurfaceGeometry` object the
values are computed via `curvature(x)`. These values are forwarded to the
JavaScript viewer for curvature based coloring.

## Development Setup

This package includes an interactive 3D visualization component powered by [surfviewjs](https://github.com/bbuchsbaum/surfviewjs). The built UMD bundle is committed so `devtools::install_github()` works without Node; Node is only needed if you want to rebuild the library yourself. (Current surfviewjs: 2.1.0)

### Prerequisites

- Node.js (v14 or higher)
- npm

### Building surfview

```bash
# Build the surfviewjs library and copy to neurosurf
make surfview

# Check build status
make status

# Clean build artifacts
make clean-surfview
```

The Makefile will:

1. Install npm dependencies in surfviewjs if needed
2. Build the library using vite
3. Copy the UMD bundle to `inst/htmlwidgets/lib/neurosurface/`
4. Automatically sync the version number to `surfwidget.yaml`

### Workflow Integration

```bash
# Before R CMD check
make r-check  # Builds surfview automatically

# Install package
make r-install  # Builds surfview automatically

# Watch for changes during development
make watch-surfview  # Requires fswatch
```

### Version Management

The version number in `surfwidget.yaml` is automatically synchronized with `surfviewjs/package.json` when running `make surfview`. To manually update:

```bash
make update-yaml
make show-version  # Verify versions are in sync
```
