---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
[![Travis-CI Build Status](https://app.travis-ci.com/bbuchsbaum/neurosurf.svg?branch=master)](https://app.travis-ci.com/bbuchsbaum/neurosurf)

# neurosurf

`neurosurf` is an R package for reading, manipulating, and visualizing
surface-based neuroimaging data represented as triangle meshes. It focuses on
surface geometry, vertex-wise data, smoothing, geodesic neighborhoods, and
both static and interactive rendering workflows.

The package is under active development.

## Installation

Install the development version from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("bbuchsbaum/neurosurf")
```

## Start here

The pkgdown site includes a few good entry points:

- [Introduction to NeuroSurf Data Structures](https://bbuchsbaum.github.io/neurosurf/articles/introduction-to-neurosurf.html)
- [Displaying Surfaces with RGL](https://bbuchsbaum.github.io/neurosurf/articles/displaying-surfaces.html)
- [Interactive Surface Visualization with surfwidget](https://bbuchsbaum.github.io/neurosurf/articles/interactive-surfaces.html)
- [Publication-quality surface figures](https://bbuchsbaum.github.io/neurosurf/articles/index.html)
- [Reference index](https://bbuchsbaum.github.io/neurosurf/reference/index.html)

## Development setup

`neurosurf` includes an interactive 3D visualization component powered by
[surfviewjs](https://github.com/bbuchsbaum/surfviewjs). The report-safe embed
bundle is committed, so package users do not need Node. Maintainers control the
source checkout at `~/code/jscode/surfviewjs`; `make surfview` rebuilds it and
records the exact source commit and artifact checksum in
`inst/htmlwidgets/lib/neurosurface/surfview.embed.commit`.

### Prerequisites

- Node.js 18 or higher
- npm

### Building surfviewjs

```bash
# Build the surfviewjs library and copy it into neurosurf
make surfview

# Verify the copied artifact and provenance marker
make verify-build

# Clean build artifacts
make clean-surfview
```

The Makefile will:

1. Build the dedicated embed bundle from the controlled `surfviewjs` checkout
2. Copy `surfview.embed.iife.js` into the R package
3. Record its source commit, package version, SHA-256, and Three.js revision
4. Verify that `surfwidget.yaml` loads that one self-contained dependency

### Version and provenance

```bash
make show-version
```

`paper-light` is the default appearance preset: a white-background,
publication-oriented combination of lighting, material, and figure-export
defaults. It is not a widget class or an interaction mode. `mode = "report"`
separately enables the compact map, anatomical-view, reset, fullscreen, and PNG
controls. The old Tweakpane control surface is deprecated and is not included
in generated reports.
