---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# neurosurf

[![r-universe](https://bbuchsbaum.r-universe.dev/badges/neurosurf)](https://bbuchsbaum.r-universe.dev/neurosurf)
[![R-CMD-check](https://github.com/bbuchsbaum/neurosurf/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bbuchsbaum/neurosurf/actions/workflows/R-CMD-check.yaml)
[Documentation](https://bbuchsbaum.github.io/neurosurf/) ·
[Reference](https://bbuchsbaum.github.io/neurosurf/reference/index.html) ·
[News](NEWS.md)

neurosurf is an R package for cortical surface data. It reads FreeSurfer,
GIFTI, and AFNI/SUMA meshes, keeps vertex-wise values attached to their
geometry, and turns them into publication figures and self-contained
interactive HTML reports.

> **Status:** Pre-release and not on CRAN. APIs may change.

## Installation

Install a prebuilt binary from r-universe:

```{r install-universe, eval = FALSE}
install.packages("neurosurf",
  repos = c("https://bbuchsbaum.r-universe.dev", "https://cloud.r-project.org")
)
```

or build the development version from GitHub (requires a C++ toolchain):

```{r install-github, eval = FALSE}
# install.packages("remotes")
remotes::install_github("bbuchsbaum/neurosurf")
```

## Quick start

Render a thresholded map on both hemispheres of the bundled fsaverage5
surface, with sulcal depth as the anatomical underlay:

```{r quickstart, message = FALSE, fig.alt = "Lateral and medial views of both fsaverage5 hemispheres with red and blue thresholded z-map clusters over lit grey sulcal shading", fig.width = 7, fig.height = 5.1, dpi = 200, out.width = "100%"}
library(neurosurf)

surf <- load_fsaverage("fsaverage5", "inflated")
white <- load_fsaverage("fsaverage5", "white")
sulc <- load_fsaverage_sulc("fsaverage5")

# A synthetic z-map: smooth clusters centred in white-surface (x, y, z) space.
# Any numeric vector with one value per vertex works.
zmap <- function(g, side) {
  cluster <- function(centre, z, width = 12) {
    z * exp(-colSums((t(coords(g)) - centre)^2) / (2 * width^2))
  }
  cluster(c(45 * side, -60, 20), 4) + cluster(c(40 * side, 20, 30), 3) +
    cluster(c(55 * side, -20, 0), -3.5) + cluster(c(8 * side, 40, 10), -3)
}

fig <- surface_figure(
  lh = surf$lh, rh = surf$rh,
  values = list(lh = zmap(white$lh, -1), rh = zmap(white$rh, 1)),
  anatomy = sulc,
  threshold = 1.5, limits = c(-4, 4), legend_title = "z"
)
plot(fig)
```

`write_surface_figure(fig, "figure.png")` saves the same figure. Rendering
runs on the CPU, so it works identically on a laptop, a cluster node, or CI,
with no OpenGL device or browser.

## What it covers

- **Read and write surfaces and data**: FreeSurfer, GIFTI, AFNI/SUMA, and NIML
  (`read_surf()`, `read_surf_geometry()`, `write_surf_data()`), plus bundled
  fsaverage5 surfaces and sulcal depth (`load_fsaverage()`,
  `load_fsaverage_sulc()`).
- **Map volumes to surfaces**: sample a volumetric image onto a mesh with
  `vol_to_surf()` or `vol_to_surf_sdf()`.
- **Analyse on the mesh**: smoothing, curvature, geodesic distances,
  neighbourhood graphs, cluster thresholding, and ROI boundaries.
- **Surface searchlights** for multivariate pattern analysis
  (`SurfaceSearchlight()`, `RandomSurfaceSearchlight()`).
- **Figures**: lit, headless, deterministic multi-view figures with a shared
  colour scale (`surface_figure()`), and interactive rgl display
  (`view_surface()`).
- **Interactive reports**: combine both hemispheres and several named maps in
  one `surface_scene()`, show it with `surfwidget()` in R Markdown or Quarto,
  or write it as a standalone HTML page with `write_surface_scene()`. The
  viewer is bundled, so reports need no CDN or network access.

Volumetric data structures come from
[neuroim2](https://github.com/bbuchsbaum/neuroim2).

## Documentation

- [Introduction to neurosurf data structures](https://bbuchsbaum.github.io/neurosurf/articles/introduction-to-neurosurf.html):
  geometry, vertex data, and the core classes.
- [Publication-quality surface figures](https://bbuchsbaum.github.io/neurosurf/articles/surface-figures.html):
  static figures for papers.
- [Bilateral interactive surface reports](https://bbuchsbaum.github.io/neurosurf/articles/interactive-surfaces.html):
  browser-based reports with several maps.
- [Displaying surfaces with rgl](https://bbuchsbaum.github.io/neurosurf/articles/displaying-surfaces.html):
  interactive exploration from an R session.
- [Function reference](https://bbuchsbaum.github.io/neurosurf/reference/index.html).

## Contributing

See [CONTRIBUTING.md](CONTRIBUTING.md), including how to rebuild the bundled
JavaScript viewer. Bug reports go to the
[issue tracker](https://github.com/bbuchsbaum/neurosurf/issues).

## License

GPL (>= 2)
