---
title: "Displaying Surfaces"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Displaying Surfaces}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Load example low desnity brain surface meshes

The `neurosurf` package includes some low density surfaces that are useful for testing and demonstrating package features. We first load left and right "inflated" and "smoothwm" surfaces that are included in the `extdata` package folder.


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rgl)
library(neuroim2)
library(neurosurf)

inflated_lh_asc <- system.file("extdata", "std.8_lh.inflated.asc", package="neurosurf")
inflated_rh_asc <- system.file("extdata", "std.8_rh.inflated.asc", package="neurosurf")

white_lh_asc <- system.file("extdata", "std.8_lh.smoothwm.asc", package="neurosurf")
white_rh_asc <- system.file("extdata", "std.8_rh.smoothwm.asc", package="neurosurf")

```


## Smoothing surface geometry

It is is often useful to smooth a surface mesh for visualization and other purposes. `neurosurf` uses the `Rvcg` package to smooth surface meshes using various smoothing approaches. Below we show smoothed surfaces using the Taubin and HCLaplace methods.


```{r, fig.width=8, fig.height=5} 

white_surf <- read_surf(white_lh_asc)
white_rh_surf <- read_surf(white_rh_asc)

open3d()
layout3d(matrix(1:4, 2,2), heights=c(1,3))
t1 <- text3d(0,0,0,"Taubin smooth, lambda=.5"); next3d()
white_surf1 <- smooth(white_surf, type="taubin", lambda=.8)
p1 <- plot(white_surf1); next3d()

t2 <- text3d(0,0,0,"HCLaplace smooth, delta=.2, iteration=5"); next3d()
white_surf2 <- smooth(white_surf, type="HCLaplace", delta=.2, iteration=5)
p2 <- plot(white_surf2)
rglwidget()
```


## Adding a color layer to a surface mesh

Here we add a colors to the surface that are a function of each surface node's `x` coordinate. When then map these values to a `rainbow` color map.

```{r, fig.width=7, fig.height=5} 

mfrow3d(1, 2, byrow = TRUE)
xvals <- white_surf1@mesh$vb[1,]
p <- plot(white_surf2, vals=xvals, cmap=rainbow(255))
next3d()
p2 <- plot(white_surf2, vals=xvals, cmap=heat.colors(255))
rglwidget()

```

## Showing an activation map overlaid on a surface mesh

We will plot surface in a row of 3. We generate a set of random values and then smooth those values along the surface to approximate a realistic activation pattern.

In the first column we display all the values in the map. Next we threshold all values between (-2,2). In the last panel we additionally add a cluster size threshold of 30 nodes.

```{r, fig.width=7, fig.height=5} 

open3d()
mfrow3d(1, 3, byrow = TRUE)
vals <- rnorm(length(nodes(white_surf2)))
surf <- NeuroSurface(white_surf2, indices=1:length(vals), data=vals)
ssurf <- smooth(surf)
p <- plot(ssurf@geometry, vals=ssurf@data, cmap=rainbow(100), irange=c(-2,2))

next3d()
comp <- conn_comp(ssurf, threshold=c(-.2,.2))
p2 <- plot(ssurf@geometry, vals=ssurf@data, cmap=rainbow(100), irange=c(-2,2), thresh=c(-.2, .2))

next3d()
csurf <- cluster_threshold(ssurf, size=30, threshold=c(-.2,.2))
p2 <- plot(csurf, cmap=rainbow(100), irange=c(-2,2), thresh=c(-.2, .2))
rglwidget()

```

## Showing two hemisperes in same scene

```{r, fig.width=6, fig.height=5} 

open3d()

curv_lh <- curvature(white_surf2)
white_rh_surf2 <- smooth(white_rh_surf, type="HCLaplace", delta=.2, iteration=5)
curv_rh <- curvature(white_rh_surf2)

p <- plot(white_surf2, bgcol=curv_cols(curv_lh), viewpoint="posterior")
p <- plot(white_rh_surf2,bgcol=curv_cols(curv_rh), viewpoint="posterior")
rglwidget()
```





