---
title: "Displaying Surfaces with RGL"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\\VignetteIndexEntry{Displaying Surfaces with RGL}
  %\\VignetteEngine{knitr::rmarkdown}
  %\\VignetteEncoding{UTF-8}
---

This vignette demonstrates how to display 3D brain surface meshes using the `rgl` plotting tools provided by the `neurosurf` package, primarily through the `plot()` method which utilizes the `view_surface()` function internally.

## Setup and Loading Data

First, we set up `knitr` options to embed `rgl` plots directly into the HTML output using WebGL and prevent standalone `rgl` windows from popping up during knitting. We then load example left and right hemisphere white matter surfaces included with the package and prepare some data (smoothed geometry, curvature, random values) for the examples.

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,        # Default figure width
  fig.height = 5,       # Default figure height
  rgl.newwindow = FALSE, # Prevent new RGL windows for chunks
  webgl = TRUE          # Use WebGL for embedding
)
rgl::setupKnitr()       # Setup rgl hook for knitr

library(rgl)
library(neuroim2)
library(neurosurf)

# Load example surfaces
white_lh_asc <- system.file("extdata", "std.8_lh.smoothwm.asc", package="neurosurf")
white_rh_asc <- system.file("extdata", "std.8_rh.smoothwm.asc", package="neurosurf")
white_lh <- read_surf(white_lh_asc)
white_rh <- read_surf(white_rh_asc)

# Prepare data for examples
white_lh_smooth <- smooth(white_lh, type="HCLaplace", delta=.2, iteration=5)
curv_lh <- curvature(white_lh_smooth)
set.seed(123) # for reproducibility
random_vals <- rnorm(length(nodes(white_lh_smooth)))
```

## Basic Surface Plotting

The simplest way to display a `SurfaceGeometry` object is using the `plot()` method. By default, it renders the surface with a light gray background. We can specify a `viewpoint`.

```{r basic-plot}
# Plot the smoothed left hemisphere from a lateral viewpoint
plot(white_lh_smooth, viewpoint="lateral")
rglwidget()
```

## Coloring Based on Curvature

Surface curvature helps distinguish gyri (outward folds) from sulci (inward folds). The `curvature()` function calculates this, and `curv_cols()` provides a simple binary color mapping (default: light gray for positive/gyri, dark gray for negative/sulci). We can pass these colors to the `bgcol` argument of `plot()` to color the surface background.

```{r curvature-plot}
# Calculate curvature colors
curv_colors <- curv_cols(curv_lh)

# Plot with curvature background from a medial viewpoint
plot(white_lh_smooth, bgcol = curv_colors, viewpoint="medial")
rglwidget()
```

## Overlaying Data Values

Often, we want to visualize data mapped onto the surface vertices (e.g., activation values, thickness). We can pass a vector of values to the `vals` argument. The `cmap` argument specifies the color map, and `irange` defines the data range to map onto the colormap. Values outside `irange` are clamped to the minimum or maximum color.

```{r data-overlay}
# Overlay random data using a rainbow colormap
# Map data range from -2 to 2 onto the colormap
plot(white_lh_smooth, vals = random_vals, cmap = rainbow(256),
     irange = c(-2, 2), viewpoint="lateral")
rglwidget()
```

## Thresholding Data Visualization

The `thresh` argument (a vector of two values, `c(lower, upper)`) can be used with `vals` to make parts of the surface transparent. Vertices where the corresponding value in `vals` is *outside* this range (i.e., less than `lower` or greater than `upper`) are rendered transparently. This is useful for focusing on specific value ranges.

```{r threshold-plot}
# Same data overlay as above, but make values between -1 and 1 transparent
plot(white_lh_smooth, vals = random_vals, cmap = rainbow(256),
     irange = c(-2, 2), thresh = c(-1, 1), viewpoint="lateral")
rglwidget()
```
*Note: Thresholding makes values *outside* the specified `thresh` range transparent. This might seem counter-intuitive if you expect it to show only values *within* the range. Be mindful of this behavior.*


## Direct Vertex Coloring

Instead of mapping data values to a colormap, you can provide a vector of specific hex color codes directly to the `vert_clrs` argument. This overrides `vals` and `cmap`. The vector length must match the number of vertices.

```{r vertex-color-plot}
# Color vertices based on their x-coordinate (e.g., red for positive x, blue for negative)
x_coords <- coords(white_lh_smooth)[, 1]
vertex_colors <- ifelse(x_coords > median(x_coords), "#FF0000", "#0000FF") # Red/Blue

plot(white_lh_smooth, vert_clrs = vertex_colors, viewpoint="ventral")
rglwidget()
```

## Controlling Transparency

The `alpha` argument controls the overall transparency of the surface, ranging from 0 (fully transparent) to 1 (fully opaque).

```{r alpha-plot}
# Plot the surface with 60% opacity (40% transparent)
plot(white_lh_smooth, vals = random_vals, cmap = heat.colors(256),
     irange = c(-2, 2), alpha = 0.6, viewpoint="posterior")
rglwidget()
```

## Adjusting Lighting and Material

The appearance of the surface is affected by lighting. The `specular` argument controls the color of specular highlights (shininess). Setting it to `"black"` creates a matte appearance.

```{r lighting-plot}
# Plot with a matte finish (no specular highlights)
plot(white_lh_smooth, vals = random_vals, cmap = topo.colors(256),
     irange = c(-2, 2), specular = "black", viewpoint="lateral")
rglwidget()
```

## Changing Viewpoints

The `viewpoint` argument can be set to common anatomical views like `"lateral"`, `"medial"`, `"ventral"`, or `"posterior"`. The function automatically selects the correct left/right version based on the surface's hemisphere information (`surf@hemi`).

```{r viewpoints-plot, fig.height=8}
# Display multiple viewpoints using rgl's layout functions
mfrow3d(2, 2, sharedMouse = TRUE)
plot(white_lh_smooth, viewpoint="lateral")
plot(white_lh_smooth, viewpoint="medial")
plot(white_lh_smooth, viewpoint="ventral")
plot(white_lh_smooth, viewpoint="posterior")
rglwidget()
```

## Displaying Two Hemispheres

You can plot multiple surfaces in the same `rgl` scene. When plotting the second surface, use `new_window = FALSE` to add it to the existing window. You might need to use the `offset` argument to position the second hemisphere correctly relative to the first.

```{r two-hemispheres-plot}
# Smooth the right hemisphere and get its curvature
white_rh_smooth <- smooth(white_rh, type="HCLaplace", delta=.2, iteration=5)
curv_rh <- curvature(white_rh_smooth)

# Plot LH with curvature background (opens the scene)
plot(white_lh_smooth, bgcol = curv_cols(curv_lh), viewpoint="lateral")

# Plot RH in the same scene, slightly offset along the x-axis
# Use new_window=FALSE to add to the current plot
plot(white_rh_smooth, bgcol = curv_cols(curv_rh), viewpoint="lateral",
     new_window = FALSE, offset = c(5, 0, 0))

# Adjust the overall view if needed (optional)
# view3d(theta = 0, phi = 0, zoom = 0.8)

rglwidget()
```

## Adding Spheres to the Surface

The `spheres` argument allows you to draw spherical markers at specified coordinates. It requires a data frame with columns `x`, `y`, `z`, and `radius`. An optional `color` column can specify colors for each sphere.

```{r spheres-plot}
# Define coordinates for some spherical markers
peak_coords <- data.frame(
  x = coords(white_lh_smooth)[c(100, 500, 1000), 1], # Example vertex coordinates
  y = coords(white_lh_smooth)[c(100, 500, 1000), 2],
  z = coords(white_lh_smooth)[c(100, 500, 1000), 3],
  radius = c(3, 4, 2.5),
  color = c("yellow", "cyan", "magenta")
)

# Plot the surface and add the spheres
plot(white_lh_smooth, viewpoint = "lateral", spheres = peak_coords)
rglwidget()
```

## Plotting Other NeuroSurface Objects

The `plot()` method also works for other classes like `NeuroSurface`, `LabeledNeuroSurface`, and `ColorMappedNeuroSurface`. These objects already contain data and potentially color mapping information. The `plot` method extracts this information and passes the appropriate arguments (like `vals`, `cmap`, `irange`, `thresh`, `vert_clrs`) to the underlying `view_surface` function.

```{r neurosurface-plot}
# Create a NeuroSurface object with the random data
nsurf <- NeuroSurface(white_lh_smooth, indices=1:length(random_vals), data=random_vals)

# Plot the NeuroSurface - uses data stored within the object
# We can still override or add parameters like cmap, irange, thresh, alpha etc.
plot(nsurf, cmap=heat.colors(128), irange=c(-2.5, 2.5), viewpoint="lateral")
rglwidget()

```


## Showing an activation map overlaid on a surface mesh

We will plot surface in a row of 3. We generate a set of random values and then smooth those values along the surface to approximate a realistic activation pattern.

In the first column we display all the values in the map. Next we threshold all values between (-2,2). In the last panel we additionally add a cluster size threshold of 30 nodes.

```{r, fig.width=7, fig.height=5} 

open3d()
mfrow3d(1, 3, byrow = TRUE)
vals <- rnorm(length(nodes(white_surf2)))
surf <- NeuroSurface(white_surf2, indices=1:length(vals), data=vals)
ssurf <- smooth(surf)
p <- plot(geometry(ssurf), vals=values(ssurf), cmap=rainbow(100), irange=c(-2,2))

next3d()
comp <- conn_comp(ssurf, threshold=c(-.2,.2))
p2 <- plot(geometry(ssurf), vals=values(ssurf), cmap=rainbow(100), irange=c(-2,2), thresh=c(-.2, .2))

next3d()
csurf <- cluster_threshold(ssurf, size=30, threshold=c(-.2,.2))
p2 <- plot(csurf, cmap=rainbow(100), irange=c(-2,2), thresh=c(-.2, .2))
rglwidget()

```

## Showing two hemisperes in same scene

```{r, fig.width=6, fig.height=5} 

open3d()

curv_lh <- curvature(white_surf2)
white_rh_surf2 <- smooth(white_rh_surf, type="HCLaplace", delta=.2, iteration=5)
curv_rh <- curvature(white_rh_surf2)

p <- plot(white_surf2, bgcol=curv_cols(curv_lh), viewpoint="posterior")
p <- plot(white_rh_surf2,bgcol=curv_cols(curv_rh), viewpoint="posterior")
rglwidget()
```

## Interactive Surface Visualization with plot_js

We can use the `plot_js` function to create an interactive 3D visualization of our surface using HTMLWidgets. This allows for a more dynamic viewing experience directly in the HTML output.

```{r, fig.width=7, fig.height=5}

# Create an interactive 3D visualization of white_surf1
surfwidget(white_surf1, width = "100%", height = "400px")
```

The widget accepts an optional `curvature` argument. If omitted it will be
derived from the geometry when `white_surf1` is a `SurfaceGeometry` object:

```{r, eval=FALSE}
surfwidget(white_surf1, curvature = curv_lh)
```

=======



