---
title: Displaying Surfaces with RGL
author: Bradley Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Displaying Surfaces with RGL}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',()=>document.body.classList.add('palette-red'));</script>

---

This vignette demonstrates how to display 3D brain surface meshes using the `rgl` plotting tools provided by the `neurosurf` package, primarily through the `plot()` method which utilizes the `view_surface()` function internally.

For interactive HTML widgets, see *Interactive Surface Visualization with surfwidget*.
For high-level, surfplot-style multi-view layouts with shared colourbars and atlas outlines, see *Surfplot-style Figures with neurosurf*.

## Setup and Loading Data

First, we set up `knitr` options to embed `rgl` plots directly into the HTML output using WebGL and prevent standalone `rgl` windows from popping up during knitting. We then load example left and right hemisphere white matter surfaces included with the package and prepare some data (smoothed geometry, curvature, random values) for the examples.

```{r setup, include = FALSE}
# Check for required packages - skip vignette build if rgl unavailable
has_rgl <- requireNamespace("rgl", quietly = TRUE)
if (!has_rgl) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Skipping vignette: rgl package not available")
}

if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,        # Default figure width
  fig.height = 5        # Default figure height
)

if (has_rgl) {
  # Use the NULL device to suppress external windows and embed widgets
  options(rgl.useNULL = TRUE, rgl.printRglwidget = TRUE)

  # Let rgl autoprint widgets into the document (>= 0.102.21)
  rgl::setupKnitr(autoprint = TRUE)

  # Additional knitr defaults for figure sizing in vignettes
  knitr::opts_chunk$set(
    dev = 'png',
    out.width = "100%",
    out.height = "400px"
  )
  options(rgl.max.lights = 2)
  library(rgl)
  options(rgl.useNULL = TRUE)
}
library(neuroim2)
library(neurosurf)

# Helper function: try snapshot_surface first, fall back to rglwidget
render_surface <- function(surfgeom, ..., width = 800, height = 600) {
  img_path <- try(snapshot_surface(surfgeom, width = width, height = height, ...), silent = TRUE)
  if (!inherits(img_path, "try-error") && length(img_path) > 0 && file.exists(img_path)) {
    knitr::include_graphics(img_path)
  } else {
    # Fallback to rglwidget
    rgl::open3d()
    view_surface(surfgeom, new_window = FALSE, ...)
    widget <- rgl::rglwidget()
    rgl::close3d()
    widget
  }
}

# Helper function: render multiple viewpoints in a grid
render_multi_view <- function(surfgeom, viewpoints = c("lateral", "medial", "ventral", "posterior"),
                              ..., width = 600, height = 450) {
  paths <- lapply(viewpoints, function(vp) {
    try(snapshot_surface(surfgeom, viewpoint = vp, width = width, height = height, ...), silent = TRUE)
  })

  valid <- sapply(paths, function(p) !inherits(p, "try-error") && length(p) > 0 && file.exists(p))

  if (all(valid)) {
    knitr::include_graphics(unlist(paths))
  } else {
    # Fallback to rglwidget with mfrow3d
    n <- length(viewpoints)
    ncol <- min(n, 2)
    nrow <- ceiling(n / ncol)
    rgl::open3d()
    rgl::mfrow3d(nrow, ncol, sharedMouse = TRUE)
    for (i in seq_along(viewpoints)) {
      if (i > 1) rgl::next3d()
      view_surface(surfgeom, viewpoint = viewpoints[i], new_window = FALSE, ...)
    }
    widget <- rgl::rglwidget()
    rgl::close3d()
    widget
  }
}

# Load example surfaces
white_lh_asc <- system.file("extdata", "std.8_lh.smoothwm.asc", package="neurosurf")
white_rh_asc <- system.file("extdata", "std.8_rh.smoothwm.asc", package="neurosurf")
white_lh <- read_surf(white_lh_asc)
white_rh <- read_surf(white_rh_asc)

# Prepare data for examples
white_lh_smooth <- smooth(white_lh, type="HCLaplace", delta=.2, iteration=5)
curv_lh <- curvature(white_lh_smooth)
set.seed(123) # for reproducibility
random_vals <- rnorm(length(nodes(white_lh_smooth)))

# Keep a base (coarse) mesh for algorithms; use a display copy we may refine
white_lh_base <- white_lh_smooth
white_lh_display <- white_lh_base
curv_lh_display <- curv_lh
random_vals_display <- random_vals
random_vals_display_smooth <- random_vals_display

```

```{r upsample-hidden, echo=FALSE, message=FALSE, warning=FALSE}
# Gently refine the std8 mesh to reduce faceting in the vignette visuals
if (requireNamespace("Rvcg", quietly = TRUE)) {
  try({
    white_lh_hi <- white_lh_base
    if (is.function(Rvcg::vcgSubdivide)) {
      white_lh_hi@mesh <- Rvcg::vcgSubdivide(white_lh_hi@mesh, type = "loop", iterations = 1, silent = TRUE)
    }
    white_lh_hi <- smooth(white_lh_hi, type = "HCLaplace", delta = 0.12, iteration = 3)
    white_lh_display <- white_lh_hi
    curv_lh_display <- curvature(white_lh_display)
    random_vals_display <- rnorm(length(nodes(white_lh_display))) # regenerate to match new vertex count
  }, silent = TRUE)
}

# Create a smoothed version of the random data to avoid speckled overlays
try({
  random_surf <- NeuroSurface(white_lh_display,
                              indices = seq_len(length(random_vals_display)),
                              data = random_vals_display)
  random_surf_sm <- smooth(random_surf)
  random_vals_display_smooth <- values(random_surf_sm)
}, silent = TRUE)

# Final safety: ensure data vectors match the current display vertex count
nv_display <- length(nodes(white_lh_display))

# Regenerate raw data only if the vertex count changed
if (length(random_vals_display) != nv_display) {
  random_vals_display <- rnorm(nv_display)
}

# (Re)compute a smoothed version aligned with the current mesh; fall back to raw data on error
random_vals_display_smooth <- tryCatch({
  rsurf <- NeuroSurface(white_lh_display,
                        indices = seq_len(nv_display),
                        data = random_vals_display)
  values(smooth(rsurf))
}, error = function(e) random_vals_display)

# Guard against length/NA issues after smoothing
if (length(random_vals_display_smooth) != nv_display ||
    any(!is.finite(random_vals_display_smooth))) {
  random_vals_display_smooth <- random_vals_display
}

# Recompute curvature for the current display mesh
curv_lh_display <- curvature(white_lh_display)
```

## Basic Surface Plotting

The simplest way to display a `SurfaceGeometry` object is using the `plot()` method. By default, it renders the surface with a light gray background. We can specify a `viewpoint`.



```{r basic-plot, results='asis'}
# Plot the smoothed left hemisphere from a lateral viewpoint
render_surface(white_lh_display, viewpoint = "lateral", lit = TRUE)
```

## Coloring Based on Curvature

Surface curvature helps distinguish gyri (outward folds) from sulci (inward folds). The `curvature()` function calculates this, and `curv_cols()` provides a simple binary color mapping (default: light gray for positive/gyri, dark gray for negative/sulci). We can pass these colors to the `bgcol` argument of `plot()` to color the surface background.

```{r curvature-plot, results='asis'}
# Calculate curvature colors
curv_colors <- curv_cols(curv_lh_display)

# Plot with curvature background from a medial viewpoint
render_surface(white_lh_display, bgcol = curv_colors, viewpoint = "medial", specular = "black")
```

## Overlaying Data Values

Often, we want to visualize data mapped onto the surface vertices (e.g., activation values, thickness). We can pass a vector of values to the `vals` argument. The `cmap` argument specifies the color map, and `irange` defines the data range to map onto the colormap. Values outside `irange` are clamped to the minimum or maximum color.

```{r data-overlay, results='asis'}
# Overlay random data using a rainbow colormap
# Map data range from -2 to 2 onto the colormap
render_surface(white_lh_display, vals = random_vals_display_smooth, cmap = rainbow(256),
               irange = c(-2, 2), thresh = NULL, viewpoint = "lateral", specular = "gray")
```

## Thresholding Data Visualization

The `thresh` argument (a vector of two values, `c(lower, upper)`) can be used with `vals` to make parts of the surface transparent. Vertices where the corresponding value in `vals` is *inside* this range (between `lower` and `upper`) are rendered transparently; values outside remain opaque. This is useful for masking out a band of values.

```{r threshold-plot, results='asis'}
# Same data overlay as above, but make values between -1 and 1 transparent
render_surface(white_lh_display, vals = random_vals_display_smooth, cmap = rainbow(256),
               irange = c(-2, 2), thresh = c(-1, 1), viewpoint = "lateral", lit = TRUE)
```

## Direct Vertex Coloring

Instead of mapping data values to a colormap, you can provide a vector of specific hex color codes directly to the `vert_clrs` argument. This overrides `vals` and `cmap`. The vector length must match the number of vertices.

```{r vertex-color-plot, results='asis'}
# Color vertices based on their x-coordinate (e.g., red for positive x, blue for negative)
x_coords <- coords(white_lh_display)[, 1]
vertex_colors <- ifelse(x_coords > median(x_coords), "#FF0000", "#0000FF") # Red/Blue

render_surface(white_lh_display, vert_clrs = vertex_colors, viewpoint = "ventral", lit = TRUE)
```

## Controlling Transparency

The `alpha` argument controls the overall transparency of the surface, ranging from 0 (fully transparent) to 1 (fully opaque).

```{r alpha-plot, results='asis'}
# Plot the surface with 60% opacity (40% transparent)
render_surface(white_lh_display, vals = random_vals_display_smooth, cmap = heat.colors(256),
               irange = c(-2, 2), alpha = 0.6, viewpoint = "posterior")
```

## Adjusting Lighting and Material

The appearance of the surface is affected by lighting. The `specular` argument controls the color of specular highlights (shininess). Setting it to `"black"` creates a matte appearance.

```{r lighting-plot, results='asis'}
# Plot with a matte finish (no specular highlights)
render_surface(white_lh_display, vals = random_vals_display_smooth, cmap = topo.colors(256),
               irange = c(-2, 2), specular = "black", viewpoint = "lateral", lit = TRUE)
```

## Snapshotting to an image (for knitr/CI)

Use \code{snapshot_surface()} to render an off-screen PNG and include it directly:

```{r snapshot-example}
img_path <- try(snapshot_surface(white_lh_display,
                                 vals = random_vals_display_smooth,
                                 cmap = viridis::viridis(256),
                                 viewpoint = "lateral",
                                 specular = "black",
                                 width = 1200, height = 900),
                silent = TRUE)

if (!inherits(img_path, "try-error") && is.character(img_path) && file.exists(img_path)) {
  knitr::include_graphics(img_path)
} else {
  cat("*(Snapshot unavailable in this build environment)*")
}
```

## Changing Viewpoints

The `viewpoint` argument can be set to common anatomical views like `"lateral"`, `"medial"`, `"ventral"`, or `"posterior"`. The function automatically selects the correct left/right version based on the surface's hemisphere information (`surf@hemi`).

```{r viewpoints-plot, results='asis', fig.height=8, out.width="50%", fig.show='hold'}
# Display multiple viewpoints
render_multi_view(white_lh_display, viewpoints = c("lateral", "medial", "ventral", "posterior"))
```

## Displaying Two Hemispheres

You can plot multiple surfaces in the same `rgl` scene. When plotting the second surface, use `new_window = FALSE` to add it to the existing window. You might need to use the `offset` argument to position the second hemisphere correctly relative to the first.

```{r two-hemispheres-plot, results='asis'}
# Smooth the right hemisphere and get its curvature
white_rh_smooth <- smooth(white_rh, type = "HCLaplace", delta = 0.2, iteration = 5)
curv_rh <- curvature(white_rh_smooth)

# Try snapshot approach for two hemispheres
img_path <- try({
  file <- tempfile(fileext = ".png")
  rgl::open3d()
  rgl::par3d(windowRect = c(0, 0, 1200, 600))
  rgl::bg3d(color = "white")

  # Plot LH with curvature background

  view_surface(white_lh_display, bgcol = curv_cols(curv_lh_display),
               viewpoint = "lateral", new_window = FALSE)
  # Plot RH offset to the right
  view_surface(white_rh_smooth, bgcol = curv_cols(curv_rh),
               viewpoint = "lateral", new_window = FALSE, offset = c(100, 0, 0))

  if (rgl::rgl.useNULL() && requireNamespace("webshot2", quietly = TRUE)) {
    rgl::snapshot3d(file, webshot = TRUE)
  } else {
    rgl::rgl.snapshot(file)
  }
  rgl::close3d()
  file
}, silent = TRUE)

if (!inherits(img_path, "try-error") && file.exists(img_path)) {
  knitr::include_graphics(img_path)
} else {
  # Fallback to rglwidget
  rgl::open3d()
  view_surface(white_lh_display, bgcol = curv_cols(curv_lh_display),
               viewpoint = "lateral", new_window = FALSE)
  view_surface(white_rh_smooth, bgcol = curv_cols(curv_rh),
               viewpoint = "lateral", new_window = FALSE, offset = c(100, 0, 0))
  rgl::rglwidget()
}
```

## Adding Spheres to the Surface

The `spheres` argument allows you to draw spherical markers at specified coordinates. It requires a data frame with columns `x`, `y`, `z`, and `radius`. An optional `color` column can specify colors for each sphere.

```{r spheres-plot, results='asis'}
# Define coordinates for some spherical markers
# Sample some vertex indices safely from available vertices
n_vertices <- nrow(coords(white_lh_display))
sample_indices <- sample(1:n_vertices, size = min(3, n_vertices))

peak_coords <- data.frame(
  x = coords(white_lh_display)[sample_indices, 1],
  y = coords(white_lh_display)[sample_indices, 2],
  z = coords(white_lh_display)[sample_indices, 3],
  radius = c(3, 4, 2.5)[1:length(sample_indices)],
  color = c("yellow", "cyan", "magenta")[1:length(sample_indices)]
)

# Plot the surface and add the spheres
render_surface(white_lh_display, viewpoint = "lateral", spheres = peak_coords)
```

## Plotting Other NeuroSurface Objects

The `plot()` method also works for other classes like `NeuroSurface`, `LabeledNeuroSurface`, and `ColorMappedNeuroSurface`. These objects already contain data and potentially color mapping information. The `plot` method extracts this information and passes the appropriate arguments (like `vals`, `cmap`, `irange`, `thresh`, `vert_clrs`) to the underlying `view_surface` function.

```{r neurosurface-plot, results='asis'}
# Create a NeuroSurface object with the random data
nsurf <- NeuroSurface(white_lh_display, indices = 1:length(random_vals_display), data = random_vals_display)

# Plot the NeuroSurface - uses data stored within the object
# We can still override or add parameters like cmap, irange, thresh, alpha etc.
render_surface(geometry(nsurf), vals = values(nsurf), cmap = heat.colors(128),
               irange = c(-2.5, 2.5), viewpoint = "lateral")
```


## Showing an activation map overlaid on a surface mesh

We will plot surface in a row of 3. We generate a set of random values and then smooth those values along the surface to approximate a realistic activation pattern.

In the first column we display all the values in the map. Next we threshold all values between (-2,2). In the last panel we additionally add a cluster size threshold of 30 nodes.

```{r activation-map, results='asis', fig.width=7, fig.height=5, out.width="33%", fig.show='hold'}
vals <- rnorm(length(nodes(white_lh_base)))
surf <- NeuroSurface(white_lh_base, indices = 1:length(vals), data = vals)
ssurf <- smooth(surf)

# Panel 1: All values
img1 <- try(snapshot_surface(geometry(ssurf), vals = values(ssurf), cmap = rainbow(100),
                              irange = c(-2, 2), width = 600, height = 450), silent = TRUE)

# Panel 2: Thresholded (values between -0.2 and 0.2 transparent)
comp <- conn_comp(ssurf, threshold = c(-0.2, 0.2))
img2 <- try(snapshot_surface(geometry(ssurf), vals = values(ssurf), cmap = rainbow(100),
                              irange = c(-2, 2), thresh = c(-0.2, 0.2), width = 600, height = 450), silent = TRUE)

# Panel 3: Cluster thresholded (minimum cluster size of 30)
csurf <- cluster_threshold(ssurf, size = 30, threshold = c(-0.2, 0.2))
img3 <- try(snapshot_surface(geometry(csurf), vals = values(csurf), cmap = rainbow(100),
                              irange = c(-2, 2), thresh = c(-0.2, 0.2), width = 600, height = 450), silent = TRUE)

# Check if all snapshots succeeded
valid <- sapply(list(img1, img2, img3), function(p) !inherits(p, "try-error") && length(p) > 0 && file.exists(p))

if (all(valid)) {
  knitr::include_graphics(c(img1, img2, img3))
} else {
  # Fallback to rglwidget
  rgl::open3d()
  rgl::mfrow3d(1, 3, byrow = TRUE)
  view_surface(geometry(ssurf), vals = values(ssurf), cmap = rainbow(100),
               irange = c(-2, 2), new_window = FALSE)
  rgl::next3d()
  view_surface(geometry(ssurf), vals = values(ssurf), cmap = rainbow(100),
               irange = c(-2, 2), thresh = c(-0.2, 0.2), new_window = FALSE)
  rgl::next3d()
  view_surface(geometry(csurf), vals = values(csurf), cmap = rainbow(100),
               irange = c(-2, 2), thresh = c(-0.2, 0.2), new_window = FALSE)
  rgl::rglwidget()
}
```

## Showing two hemispheres in same scene

```{r two-hemi-posterior, results='asis', fig.width=6, fig.height=5}
# Two hemispheres shown from posterior viewpoint
img_path <- try({
  file <- tempfile(fileext = ".png")
  rgl::open3d()
  rgl::par3d(windowRect = c(0, 0, 1200, 600))
  rgl::bg3d(color = "white")

  view_surface(white_lh_display, bgcol = curv_cols(curv_lh_display),
               viewpoint = "posterior", new_window = FALSE)
  view_surface(white_rh_smooth, bgcol = curv_cols(curv_rh),
               viewpoint = "posterior", new_window = FALSE, offset = c(100, 0, 0))

  if (rgl::rgl.useNULL() && requireNamespace("webshot2", quietly = TRUE)) {
    rgl::snapshot3d(file, webshot = TRUE)
  } else {
    rgl::rgl.snapshot(file)
  }
  rgl::close3d()
  file
}, silent = TRUE)

if (!inherits(img_path, "try-error") && file.exists(img_path)) {
  knitr::include_graphics(img_path)
} else {
  # Fallback to rglwidget
  rgl::open3d()
  view_surface(white_lh_display, bgcol = curv_cols(curv_lh_display),
               viewpoint = "posterior", new_window = FALSE)
  view_surface(white_rh_smooth, bgcol = curv_cols(curv_rh),
               viewpoint = "posterior", new_window = FALSE, offset = c(100, 0, 0))
  rgl::rglwidget()
}
```


## Next Steps

For **interactive 3D visualization** with `surfwidget()`, see `vignette("interactive-surfaces")`.

For **publication-quality multi-view figures**, see `vignette("surfplot-style-figures")`.
