---
title: "Build bilateral interactive surface reports"
author: Bradley Buchsbaum
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Build bilateral interactive surface reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

An HTML report often needs one interactive figure for both hemispheres and
several result maps. Repeating a widget for every map wastes space and sends
the same mesh to the browser many times. `surface_scene()` describes the
geometries, maps, labels, and fallback once; `surfwidget()` displays that scene
in R Markdown or Quarto; and `write_surface_scene()` writes the same scene as
an ordinary web page.

This article builds one bilateral viewer with two selectable maps. The example
uses package data, runs without a network connection, and keeps the scientific
transformation in R.

```{r setup, include=FALSE}
Sys.setenv(RGL_USE_NULL = "TRUE")
options(rgl.useNULL = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  fig.height = 6
)

library(neurosurf)

left_path <- system.file(
  "extdata", "std.8_lh.smoothwm.asc", package = "neurosurf"
)
right_path <- system.file(
  "extdata", "std.8_rh.smoothwm.asc", package = "neurosurf"
)
left_geometry <- read_surf(left_path)
right_geometry <- read_surf(right_path)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## What does a scene contain?

A `SurfaceScene` has one `SurfaceGeometry` per hemisphere and one or more named
scalar layers. Each bilateral layer supplies a left and right value vector
whose lengths match the corresponding geometry. A scene also records the
initial map, legend metadata, provenance, alternative text, and fallback text.

The example surfaces are small meshes shipped with neurosurf:

```{r load-geometries, eval=FALSE}
left_geometry <- read_surf(system.file(
  "extdata", "std.8_lh.smoothwm.asc", package = "neurosurf"
))
right_geometry <- read_surf(system.file(
  "extdata", "std.8_rh.smoothwm.asc", package = "neurosurf"
))
```

The browser renders the values that R supplies. Perform thresholding, capping,
atlas projection, and other scientific transformations before constructing a
layer. Here we create a deterministic coordinate-based demonstration map and
replace sub-threshold values with `NA` in R:

```{r prepare-maps}
scaled_height <- function(geometry) {
  value <- coords(geometry)[, 3]
  3 * (value - mean(value)) / max(abs(value - mean(value)))
}

effect <- list(
  left = scaled_height(left_geometry),
  right = scaled_height(right_geometry)
)
effect <- lapply(effect, function(x) replace(x, abs(x) < 1.25, NA_real_))
```

The second map is a simple two-region label used to demonstrate map switching:

```{r prepare-regions}
region <- list(
  left = as.numeric(coords(left_geometry)[, 2] > 0),
  right = as.numeric(coords(right_geometry)[, 2] > 0)
)
```

```{r validate-inputs, include=FALSE}
stopifnot(
  length(effect$left) == nrow(coords(left_geometry)),
  length(effect$right) == nrow(coords(right_geometry)),
  any(is.finite(effect$left)),
  any(is.finite(effect$right)),
  all(region$left %in% c(0, 1)),
  all(region$right %in% c(0, 1))
)
```

## How do you build one bilateral, multi-map viewer?

Create each named map with `surface_layer()`, then put both geometries and all
layers in one scene. `selected_layer` chooses the map shown when the viewer
opens.

```{r build-scene}
scene <- surface_scene(
  left = left_geometry,
  right = right_geometry,
  layers = list(
    surface_layer(
      "effect", effect,
      colormap = c("#2166ac", "#f7f7f7", "#b2182b"),
      limits = c(-3, 3), units = "z",
      legend = list(title = "Thresholded effect")
    ),
    surface_layer(
      "region", region,
      colormap = c("#d9d9d9", "#238b45"),
      limits = c(0, 1), units = "class",
      legend = list(title = "Demonstration region")
    )
  ),
  selected_layer = "effect",
  metadata = list(subject = "example"),
  provenance = list(source = "neurosurf package data"),
  fallback = "Static bilateral surface figure with effect and region maps.",
  alt_text = paste(
    "Bilateral cortical surfaces with selectable thresholded effect",
    "and demonstration region maps."
  ),
  preset = "paper-light",
  mode = "report"
)
```

```{r validate-scene, include=FALSE}
stopifnot(
  methods::is(scene, "SurfaceScene"),
  identical(names(scene@geometries), c("left", "right")),
  identical(names(scene@layers), c("effect", "region")),
  identical(scene@selected_layer, "effect")
)
```

Pass the scene to `surfwidget()` once. The map selector changes the active
values and legend without constructing another widget or another WebGL
context.

```{r render-scene}
surfwidget(scene, width = "100%", height = "520px")
```

The report toolbar provides the authored map selector, coordinated anatomical
views, reset, and PNG export. Drag to rotate, use the wheel or a pinch gesture
to zoom, and right-drag to pan.

## Which option controls appearance and behavior?

`paper-light` and `report` name different contracts:

| Setting | Controls | What it changes |
|---|---|---|
| `preset = "paper-light"` | Appearance | Light background, restrained lighting and material values, labels, and PNG defaults. |
| `mode = "report"` | Behavior | Lazy mounting, the compact report toolbar, deterministic initial framing, and fallback/print behavior. |
| `mode = "viewer"` | Behavior | A bare rotatable viewer without the report toolbar. |

Tweakpane is deprecated. The report runtime does not load it or contact a CDN.
Legacy `showControls` and `controlType` configuration entries warn during the
migration; use the scene `mode` instead.

## How is geometry shared across maps?

`surface_scene_manifest()` converts numeric arrays to typed binary assets and
addresses each asset by its SHA-256 digest. The scene refers to each hemisphere
geometry once. Each additional full bilateral map contributes two value
arrays, not another copy of the vertices or faces.

```{r inspect-assets}
manifest <- surface_scene_manifest(scene, asset_mode = "inline")
roles <- vapply(manifest$assets, `[[`, character(1), "role")
bytes <- vapply(manifest$assets, `[[`, numeric(1), "byteLength")

data.frame(
  content = c("geometry", "map values"),
  assets = c(sum(roles %in% c("vertices", "faces", "curvature")),
             sum(roles == "values")),
  bytes = c(sum(bytes[roles %in% c("vertices", "faces", "curvature")]),
            sum(bytes[roles == "values"]))
)
```

```{r validate-assets, include=FALSE}
stopifnot(
  length(manifest$geometries) == 2L,
  length(manifest$layers) == 2L,
  sum(roles == "vertices") == 2L,
  sum(roles == "values") == 4L
)
```

Missing map values remain IEEE `NaN` values in the binary payload. The browser
verifies every adjacent asset's byte length and checksum before constructing
the scene.

## How do you write an offline report?

Use `write_surface_scene()` when the viewer must live outside an R Markdown or
Quarto document. Both output modes bundle the compatible surfviewjs runtime
locally and make no network requests.

```{r write-reports, eval=FALSE}
# index.html, the runtime, and content-addressed .bin files
write_surface_scene(scene, "surface-report", self_contained = FALSE)

# one index.html with the runtime and typed arrays inlined
write_surface_scene(scene, "surface-report-inline", self_contained = TRUE)
```

Choose adjacent assets for a site or a report directory: the browser can cache
the runtime and SHA-addressed arrays, and repeated assets are deduplicated.
Choose a self-contained file when a single portable HTML file matters more than
file size or cross-page caching.

```{r validate-report-writers, include=FALSE}
adjacent_dir <- tempfile("neurosurf-adjacent-")
inline_dir <- tempfile("neurosurf-inline-")
adjacent_html <- write_surface_scene(
  scene, adjacent_dir, self_contained = FALSE
)
inline_html <- write_surface_scene(scene, inline_dir, self_contained = TRUE)
stopifnot(
  file.exists(adjacent_html),
  file.exists(inline_html),
  file.exists(file.path(adjacent_dir, "surfview.embed.iife.js")),
  length(list.files(adjacent_dir, pattern = "\\.bin$")) > 0L
)
```

## How do you mount the scene on a plain web page?

The HTML writer produces a minimal page that uses the public
`surfview.mountSurfView()` API. You can customize its generated `index.html` or
use the same lifecycle in another page. In this example, `manifest` is a
`surfview.scene.v1` object produced by `surface_scene_manifest()`:

```html
<div id="surface-report"></div>
<script src="./surfview.embed.iife.js"></script>
<script>
  const handle = surfview.mountSurfView(
    document.getElementById("surface-report"),
    manifest,
    {
      lazy: true,
      preset: "paper-light",
      controls: true,
      baseUrl: document.baseURI,
      bilateralGroup: {
        id: "bilateral",
        leftSurfaceId: "left",
        rightSurfaceId: "right"
      }
    }
  );

  handle.ready.then(() => handle.setView("lateral"));
  window.addEventListener("pagehide", () => handle.dispose(), { once: true });
</script>
```

A manifest with both hemispheres must name a `bilateralGroup` that pairs the
left and right surface ids; without it the report runtime refuses to mount.
`surfwidget()` and `write_surface_scene()` supply this option automatically.

The returned handle also provides `selectLayer()`, `resize()`, `exportPNG()`,
and `dispose()`. Call `dispose()` when an application removes or replaces the
viewer so it can release observers, listeners, animation frames, the WebGL
context, and GPU resources.

## What happens when interaction is unavailable?

`fallback` is required plain text. The widget and standalone writer expose it
when JavaScript is disabled or WebGL initialization fails, and print CSS shows
it instead of a blank canvas. `alt_text` labels the interactive figure for
assistive technology.

The fallback text is not a bitmap renderer. If a report needs a visual print,
PDF, or archival representation, build a static figure from the same R values
with `surface_figure()`; see `vignette("surface-figures")`.

## When should you use the legacy shortcuts?

`surfwidget()` still accepts `SurfaceGeometry`, `NeuroSurface`,
`ColorMappedNeuroSurface`, and `VertexColoredNeuroSurface`. Those methods adapt
a single surface to a `SurfaceScene` and are useful for quick exploration:

```{r legacy-shortcut, eval=FALSE}
values <- scaled_height(left_geometry)
surface <- NeuroSurface(left_geometry, nodes(left_geometry), values)
surfwidget(surface, irange = c(-3, 3))
```

Use an explicit `SurfaceScene` for bilateral figures, multiple named maps,
portable assets, authored fallback text, provenance, or report behavior.

## What should you read next?

- `vignette("surface-figures")` builds static, publication-quality multi-view
  figures with `surface_figure()`.
- `vignette("displaying-surfaces")` covers RGL-based 3D rendering, curvature
  shading, and PNG snapshots.
- `vignette("introduction-to-neurosurf")` introduces `SurfaceGeometry`,
  `NeuroSurface`, and the package's other core data structures.
