---
title: "Interactive Surface Visualization with surfwidget"
author: "Bradley Buchsbaum"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive Surface Visualization with surfwidget}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.height = 6
)

# Ensure htmlwidgets dependencies are properly loaded
library(htmlwidgets)
library(neurosurf)

# Load example data
white_lh_asc <- system.file("extdata", "std.8_lh.smoothwm.asc", package="neurosurf")
white_lh <- read_surf(white_lh_asc)
white_lh_smooth <- smooth(white_lh, type="HCLaplace", delta=.2, iteration=5)

# Prepare sample data
set.seed(123)
random_vals <- rnorm(length(nodes(white_lh_smooth)))
```

## Introduction

The `surfwidget()` function provides interactive 3D visualization of brain surfaces using HTML widgets. This creates web-based viewers with real-time rotation, zooming, and interactive exploration capabilities.

## System Check

Let's first verify that the widget system is working properly:

```{r system-check}
# Check rendering context
cat("Knitting context:", if(isTRUE(getOption('knitr.in.progress'))) "vignette" else "interactive", "\n")
cat("htmlwidgets version:", as.character(packageVersion("htmlwidgets")), "\n")
cat("neurosurf version:", as.character(packageVersion("neurosurf")), "\n")

# Test basic widget creation
tryCatch({
  test_widget <- surfwidget(white_lh_smooth, width = "300px", height = "200px")
  if (inherits(test_widget, "htmlwidget")) {
    cat("✓ surfwidget creation successful\n")
    cat("✓ Widget class:", paste(class(test_widget), collapse = " "), "\n")
  } else {
    cat("✗ surfwidget creation failed - wrong class\n")
  }
}, error = function(e) {
  cat("✗ surfwidget creation failed:", e$message, "\n")
})
```

## Basic Surface Display

The simplest usage is to pass a `SurfaceGeometry` object directly:

```{r basic-widget, results='asis'}
tryCatch({
  # Create basic interactive surface
  basic_widget <- surfwidget(white_lh_smooth, width = "100%", height = "400px")
  
  if (inherits(basic_widget, "htmlwidget")) {
    cat("**Interactive surface created successfully**\n\n")
    basic_widget
  } else {
    cat("**Widget creation failed - not an htmlwidget**\n\n")
  }
}, error = function(e) {
  cat("**Error creating widget:**", e$message, "\n\n")
  cat("*Note: If you're viewing this vignette as HTML, the interactive widget may not display properly due to JavaScript loading issues in the vignette context. Try running the same code in an interactive R session.*\n\n")
})
```

## Data Visualization

Display data mapped onto surface vertices using a `NeuroSurface` object:

```{r data-widget, results='asis'}
tryCatch({
  # Create NeuroSurface with data
  nsurf <- NeuroSurface(white_lh_smooth, indices=1:length(random_vals), data=random_vals)
  
  # Create widget with data overlay
  data_widget <- surfwidget(nsurf, 
                           cmap = heat.colors(256), 
                           irange = c(-2, 2),
                           width = "100%", 
                           height = "400px")
  
  if (inherits(data_widget, "htmlwidget")) {
    cat("**Data overlay widget created successfully**\n\n")
    data_widget
  } else {
    cat("**Data widget creation failed**\n\n")
  }
}, error = function(e) {
  cat("**Error creating data widget:**", e$message, "\n\n")
})
```

## Advanced Configuration

Fine-tune the appearance with custom configuration:

```{r advanced-widget, results='asis'}
tryCatch({
  # Create ColorMappedNeuroSurface with threshold
  color_mapped_surf <- ColorMappedNeuroSurface(
    white_lh_smooth, 
    indices = 1:length(random_vals), 
    data = random_vals,
    cmap = rainbow(256),
    irange = c(-2.5, 2.5),
    thresh = c(-1, 1)
  )
  
  # Advanced configuration
  advanced_config <- list(
    shininess = 60,
    specularColor = "#ffffff",
    ambientLightColor = "#606060",
    directionalLightColor = "#ffffff",
    directionalLightIntensity = 0.7,
    flatShading = FALSE
  )
  
  advanced_widget <- surfwidget(color_mapped_surf, 
                               config = advanced_config, 
                               alpha = 0.9,
                               width = "100%", 
                               height = "450px")
  
  if (inherits(advanced_widget, "htmlwidget")) {
    cat("**Advanced configuration widget created successfully**\n\n")
    advanced_widget
  } else {
    cat("**Advanced widget creation failed**\n\n")
  }
}, error = function(e) {
  cat("**Error creating advanced widget:**", e$message, "\n\n")
})
```

## Vertex-Colored Surfaces

Create surfaces with direct vertex coloring:

```{r vertex-widget, results='asis'}
tryCatch({
  # Create vertex colors based on coordinates
  x_coords <- coords(white_lh_smooth)[, 1]
  vertex_colors <- ifelse(x_coords > median(x_coords), "#FF6B6B", "#4ECDC4")
  
  # Create VertexColoredNeuroSurface
  vertex_surf <- VertexColoredNeuroSurface(
    geometry = white_lh_smooth, 
    indices = 1:length(vertex_colors), 
    colors = vertex_colors
  )
  
  # Custom material settings
  material_config <- list(
    shininess = 20,
    specularColor = "#333333",
    ambientLightColor = "#404040"
  )
  
  vertex_widget <- surfwidget(vertex_surf, 
                             config = material_config, 
                             alpha = 0.8,
                             width = "100%", 
                             height = "400px")
  
  if (inherits(vertex_widget, "htmlwidget")) {
    cat("**Vertex-colored widget created successfully**\n\n")
    vertex_widget
  } else {
    cat("**Vertex widget creation failed**\n\n")
  }
}, error = function(e) {
  cat("**Error creating vertex widget:**", e$message, "\n\n")
})
```

## Interactive Controls

When the widgets display properly, they include several interactive controls:

- **Rotation**: Click and drag to rotate the surface
- **Zoom**: Mouse wheel or pinch gestures to zoom in/out
- **Pan**: Right-click and drag to pan the view
- **Control Panel**: Adjust lighting, materials, color mapping, and viewpoints
  - **Lighting**: Ambient and directional light colors and intensity
  - **Material**: Surface shininess (metalness/roughness)
  - **Camera**: Reset camera position and orientation
  - **Color Map**: Real-time adjustment of intensity and threshold ranges
  - **Viewpoint**: Switch between anatomical views (lateral, medial, ventral, posterior)
  - **SSAO**: Screen-space ambient occlusion for enhanced depth perception

## Troubleshooting

**Widget Creation vs. Display:**

The widgets should create successfully (as shown in the system check above), but may not display properly in vignette HTML output due to JavaScript module loading timing issues.

**Testing in Interactive R:**

All `surfwidget()` examples should work perfectly when run in an interactive R console:

```{r eval=FALSE}
# This should work in interactive R:
library(neurosurf)
white_lh_asc <- system.file("extdata", "std.8_lh.smoothwm.asc", package="neurosurf")
white_lh <- read_surf(white_lh_asc)
white_lh_smooth <- smooth(white_lh, type="HCLaplace", delta=.2, iteration=5)

# Basic widget
surfwidget(white_lh_smooth, width = "100%", height = "400px")

# With data
random_vals <- rnorm(length(nodes(white_lh_smooth)))
nsurf <- NeuroSurface(white_lh_smooth, indices=1:length(random_vals), data=random_vals)
surfwidget(nsurf, cmap = heat.colors(256), irange = c(-2, 2))
```

**Alternative: Save to File**

For guaranteed display, save widgets to standalone HTML files:

```{r eval=FALSE}
widget <- surfwidget(white_lh_smooth, width = "100%", height = "400px")
htmlwidgets::saveWidget(widget, "interactive_surface.html", selfcontained = TRUE)
# Open the resulting HTML file in your browser
```

## Advantages over Static Plots

Interactive widgets offer several advantages:

- **Real-time interaction**: Rotate, zoom, and pan without re-rendering
- **Dynamic controls**: Adjust visualization parameters interactively
- **Web-based**: Works in HTML documents, Shiny apps, and R Markdown
- **Touch support**: Works on tablets and mobile devices
- **High-quality rendering**: Uses WebGL for smooth, hardware-accelerated graphics
- **Data exploration**: Click on vertices to see coordinate and data values
- **Multiple viewpoints**: Easily switch between anatomical orientations

These interactive capabilities make `surfwidget()` ideal for exploratory data analysis and interactive presentations. 