---
title: "Reproducing rgl WebGL lighting uniform error"
output: html_document
vignette: >
  %\VignetteIndexEntry{rgl WebGL lighting uniform error}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates a minimal example that triggers the WebGL error
in a browser console when viewing an `rglwidget`:

```
WebGL: INVALID_VALUE: uniform3fv: no array
```

It does **not** depend on `neurosurf`; it uses only base **rgl** objects. The
error occurs when the shader expects more lights (`maxLights`) than are present
in the subscene, so lighting uniform arrays are too short in
`rglwidgetClass.doLighting`.

```{r setup, include=FALSE}
options(rgl.useNULL = TRUE)
library(rgl)
```

## Minimal reproducible example

The error shows up when the compiled shader expects more lights (default
`maxLights = 8`) than are actually present in the subscene. The sequence below
forces that mismatch by removing all lights before drawing a lit, specular
mesh.

```{r}
# Start clean, then remove all lights (shader still expects up to 8 by default)
clear3d(type = "all")
clear3d(type = "lights")   # zero lights in the subscene

# Vertex-colored mesh with specular highlights
sph <- subdivision3d(icosahedron3d(), depth = 3)
cols <- colorRampPalette(c('#444444', '#dddddd'))(ncol(sph$vb))
shade3d(sph, col = cols, meshColor = "vertices", specular = "white", lit = TRUE)

# Embed widget (open DevTools -> Console to see INVALID_VALUE)
rglwidget()
```

### What to look for

Open the rendered HTML in a browser with DevTools open. In Chrome and Firefox
on macOS (rgl 1.3.24) this produces:

```
WebGL: INVALID_VALUE: uniform3fv: no array
    at rglwidgetClass.doLighting (draw.src.js:275)
```

The object still appears but may render without proper lighting. Adding back
lights to match `maxLights` or reducing `maxLights` to the actual number of
lights avoids the error (e.g., `par3d(maxLights = 1)`).

### Temporary workaround

For knitting vignettes, set `specular = "black"` or `lit = FALSE` to bypass the
specular uniform upload while awaiting an upstream fix.
