---
title: Publication-quality continuous surface figures
author: neurosurf authors
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Publication-quality continuous surface figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (
  requireNamespace("systemfonts", quietly = TRUE) &&
  requireNamespace("albersdown", quietly = TRUE) &&
  "albers_register_fonts" %in% getNamespaceExports("albersdown")
) {
  getExportedValue("albersdown", "albers_register_fonts")()
}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(
    albersdown::theme_albers(
      family = params$family,
      preset = params$preset
    )
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)

library(neurosurf)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

You have a continuous value at every cortical vertex--for example a *t*, *z*,
thickness, or connectivity map--and need a static figure that remains
scientifically interpretable after thresholding. The important distinction is
that the value is a **field on a triangle mesh**, not a parcel label. A polished
figure should therefore interpolate the scalar within each triangle, test
visibility with a real depth buffer, and add anatomy without drawing atlas or
occlusion lines over the result.

Two functions cover this task. `surface_figure()` renders both hemispheres
from several views and returns one finished figure with a shared colour bar;
`write_surface_figure()` writes it to PNG. Underneath it,
`render_surface_rgba()` is the deterministic, headless single-panel primitive:
it returns an RGBA raster plus diagnostic buffers and provenance, and
`write_surface_rgba()` writes one panel to PNG. Both run identically in CI,
on a cluster, and on a laptop, with no OpenGL and no browser.

## What are the inputs?

You need a display geometry and one numeric value per vertex. For anatomical
context, use a curvature or sulcal-depth metric from a corresponding white,
pial, or midthickness surface rather than computing curvature on the inflated
display shape.

The package includes matching decimated fsaverage surfaces, which keep this
example fast and offline.

```{r load-surfaces}
inflated <- load_fsaverage_std8("inflated")

read_geom <- function(name) {
  read_surf_geometry(system.file("extdata", name, package = "neurosurf"))
}
white_lh <- read_geom("std.8_lh.white.asc")
white_rh <- read_geom("std.8_rh.white.asc")
```

Here a smooth coordinate-based field stands in for your measured statistic.
The renderer treats it exactly like any other vertex-wise numeric vector.

```{r make-statistic}
example_statistic <- function(geometry) {
  xyz <- coords(geometry)
  value <- sin(xyz[, 2] / 18) + cos(xyz[, 3] / 14) +
    0.7 * sin(xyz[, 1] / 12)
  as.numeric(scale(value))
}

stat_lh <- example_statistic(inflated$lh)
stat_rh <- example_statistic(inflated$rh)
```

```{r validate-inputs, include=FALSE}
curv_lh <- curvature(white_lh)
curv_rh <- curvature(white_rh)
stopifnot(
  length(stat_lh) == length(curv_lh),
  length(stat_rh) == length(curv_rh),
  all(is.finite(stat_lh)),
  all(is.finite(stat_rh)),
  stats::sd(stat_lh) > 0,
  stats::sd(stat_rh) > 0
)
```

## How do you render one thresholded view?

Choose the threshold and colour limits as scientific inputs. In this example,
the colour scale stays fixed at `[-2.5, 2.5]`, while an opacity ramp softens
only the visual transition immediately above `|z| = 1`.

```{r render-one-view}
lh_lateral <- render_surface_rgba(
  inflated$lh,
  vertex_values = stat_lh,
  anatomy_metric = curv_lh,
  camera = "lateral",
  threshold = 1,
  tail = "two_sided",
  limits = c(-2.5, 2.5),
  alpha_ramp = 0.25,
  antialias = 2
)
```

```{r validate-one-view, include=FALSE}
active <- as.integer(lh_lateral$overlay_alpha) > 0
stopifnot(
  any(lh_lateral$coverage),
  any(active),
  any(lh_lateral$coverage & !active),
  identical(lh_lateral$provenance$backend, "cpu_barycentric"),
  identical(lh_lateral$provenance$limits, c(-2.5, 2.5))
)
```

```{r show-one-view, echo=FALSE, out.width="80%"}
one_view_file <- knitr::fig_path("-cpu-lh-lateral.png")
dir.create(dirname(one_view_file), recursive = TRUE, showWarnings = FALSE)
write_surface_rgba(lh_lateral, one_view_file)
knitr::include_graphics(one_view_file)
```

The colour at each covered sample comes from the barycentrically interpolated
scalar. Thresholding and palette mapping happen after that interpolation, so a
threshold boundary may cross a triangle instead of jumping from one mesh edge
to the next. The per-sample z-buffer independently decides which folded surface
fragment is visible.

## How do you build the four canonical views?

`surface_figure()` renders every hemisphere-view combination with one shared
threshold, colour limit, and palette, and adds the colour bar. The camera
contract is explicit: `camera_mode = "canonical"` is strict orthographic
output, while `"presentation"` adds a small declared obliquity.

```{r render-four-views}
figure <- surface_figure(
  lh = inflated$lh,
  rh = inflated$rh,
  values = list(lh = stat_lh, rh = stat_rh),
  anatomy = list(lh = curv_lh, rh = curv_rh),
  views = c("lateral", "medial"),
  threshold = 1,
  tail = "two_sided",
  limits = c(-2.5, 2.5),
  alpha_ramp = 0.25,
  legend_title = "z",
  panel_width = 720,
  panel_height = 450,
  antialias = 2
)
```

```{r validate-four-views, include=FALSE}
stopifnot(
  length(figure$panels) == 4L,
  all(vapply(figure$panels, function(x) any(x$coverage), logical(1))),
  all(vapply(figure$panels, function(x) any(as.integer(x$overlay_alpha) > 0),
             logical(1)))
)
```

```{r show-four-views, echo=FALSE, out.width="100%"}
figure_file <- knitr::fig_path("-cpu-four-views.png")
dir.create(dirname(figure_file), recursive = TRUE, showWarnings = FALSE)
write_surface_figure(figure, figure_file)
knitr::include_graphics(figure_file)
```

In an interactive session, `plot(figure)` draws the same figure on the
current graphics device.

The lateral and medial views differ because each panel rasterizes the visible
triangles from its own camera. Internal sulcal occlusion edges are not drawn as
outlines; `outer_contour = TRUE` marks only cortex touching background that is
connected to the image exterior.

## Where do the medial wall and cortex mask enter?

A real cortex mask is a separate domain object, not a consequence of parcel
value zero. Pass one logical value per vertex. Triangles touching a masked
vertex never receive overlay colour. `medial_wall` then controls whether the
excluded domain is quietly shaded, omitted, or outlined.

```{r cortex-mask-contract, eval=FALSE}
masked_panel <- render_surface_rgba(
  inflated$lh,
  vertex_values = statistic,
  anatomy_metric = white_surface_curvature,
  cortex_mask = cortex_label,
  medial_wall = "shade"
)
```

Atlas-aware callers such as `neuroatlas::plot_brain()` can resolve the mask and
its provenance from an atlas annotation. At this low level, neurosurf requires
the caller to supply the domain explicitly rather than guessing that an atlas
label of zero always means medial wall.

## How should a volume become a continuous surface field?

For a continuous image, use explicit trilinear sampling through the cortical
ribbon. Interpolation, aggregation across depth, and tangential smoothing are
separate choices. The following contract samples five depths, averages them,
and applies no surface smoothing:

```{r ribbon-projection, eval=FALSE}
mapped <- vol_to_surf(
  white_surface,
  pial_surface,
  statistic_volume,
  fun = "avg",
  sampling = "thickness",
  interpolation = "linear",
  depth = seq(0.1, 0.9, length.out = 5),
  aggregate = "mean",
  surface_smooth_fwhm = 0
)
```

Use `interpolation = "nearest"` for discrete nearest-voxel sampling. The
historical Gaussian KNN behaviour remains available as
`interpolation = "legacy"`. Categorical `aggregate = "mode"` is intentionally
incompatible with linear interpolation.

## Which function draws what you need?

Start from the output you want:

| You want | Use |
|---|---|
| A finished static figure: views by hemispheres, one colour bar | `surface_figure()`, then `write_surface_figure()` |
| One panel, or a custom layout you compose yourself | `render_surface_rgba()`, then `write_surface_rgba()` |
| To rotate and inspect a map on your desktop during analysis | `view_surface()` |
| An interactive figure in an HTML report or Shiny app | `surfwidget()`; see `vignette("interactive-surfaces")` |

The first two run headlessly and deterministically; the last two are
interactive sessions.

Two companion packages build on the same renderer, so their pixels agree with
the figures here: `neuroatlas::plot_brain()` adds atlas-derived masks, labels,
and orientation marks, and `neuromosaic::surf_montage()` starts from a
statistic volume rather than vertex values. Atlas outlines belong in figures
about the atlas; the default continuous-statistic view omits them.

## Next steps

- `vignette("displaying-surfaces")` covers lower-level RGL rendering and local
  snapshots.
- `vignette("interactive-surfaces")` builds interactive bilateral report
  widgets with `surfwidget()`.
- `vignette("introduction-to-neurosurf")` introduces `SurfaceGeometry`,
  `NeuroSurface`, and related data structures.
- `?surface_figure`, `?render_surface_rgba`, `?surface_threshold_segments`,
  and `?vol_to_surf` document the complete computational contracts.
