---
title: Surfplot-style Figures with neurosurf
author: neurosurf authors
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surfplot-style Figures with neurosurf}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
params:
  family: red
css: albers.css
resource_files:
- albers.css
- albers.js
includes:
  in_header: |-
    <script src="albers.js"></script>
    <script>document.addEventListener('DOMContentLoaded',()=>document.body.classList.add('palette-red'));</script>

---

This vignette demonstrates how to go from fsaverage surfaces and Schaefer
parcellation labels to a multi-view, publication-ready surface figure using
neurosurf's high-level plotting API.

We will:

- load bundled fsaverage std.8 surfaces;
- load a Schaefer 200-parcel atlas projected to MNI152;
- map parcel means to the surface; and
- add atlas outlines and colourbars to produce a surfplot-style figure.

```{r setup, include=FALSE}
# Prefer rgl's NULL device for non-interactive builds (avoids headless crashes)
if (!interactive()) {
  options(rgl.useNULL = TRUE)
}

# Check for required packages - skip vignette build if rgl unavailable
has_rgl <- requireNamespace("rgl", quietly = TRUE)
if (!has_rgl) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Skipping vignette: rgl package not available")
}

if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(params$family))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

library(neurosurf)
library(neuroim2)
library(ggplot2) # used only for palette ideas
```

## Load fsaverage std.8 surfaces and Schaefer atlas

The package ships with decimated fsaverage surfaces (`std.8`). We will use
white and pial surfaces to project the Schaefer atlas from MNI152 volume
space to the surface, and inflated surfaces for visualization.

```{r load-surfaces}
# Geometry for plotting
fs_infl <- load_fsaverage_std8("inflated")
lh <- fs_infl$lh
rh <- fs_infl$rh

# White / pial surfaces for volume-to-surface mapping
lh_white <- read_surf_geometry(
  system.file("extdata", "std.8_lh.white.asc", package = "neurosurf")
)
lh_pial <- read_surf_geometry(
  system.file("extdata", "std.8_lh.pial.asc", package = "neurosurf")
)
rh_white <- read_surf_geometry(
  system.file("extdata", "std.8_rh.white.asc", package = "neurosurf")
)
rh_pial <- read_surf_geometry(
  system.file("extdata", "std.8_rh.pial.asc", package = "neurosurf")
)
```

For this example we assume that the Schaefer 200-parcel, 7-network atlas in
MNI152 space is available under `inst/extdata`. We use neuroim2 to load the
volume and `vol_to_surf()` to map parcel labels to the surface vertices.

```{r load-schaefer}
atlas_path <- system.file(
  "extdata",
  "Schaefer2018_200Parcels_7Networks_order_FSLMNI152_1mm.nii.gz",
  package = "neurosurf"
)

if (atlas_path != "") {
  atlas_img <- neuroim2::read_vol(atlas_path)

  # Project volume labels to surface vertices (mode of nearby voxels)
  lh_labels <- vol_to_surf(lh_white, lh_pial, atlas_img, fun = "mode")
  rh_labels <- vol_to_surf(rh_white, rh_pial, atlas_img, fun = "mode")
} else {
  warning("Schaefer atlas file not found in extdata; using placeholder zeros for vignette build.")
  lh_labels <- NeuroSurface(lh_white, indices = seq_len(length(nodes(lh_white))),
                            data = integer(length(nodes(lh_white))))
  rh_labels <- NeuroSurface(rh_white, indices = seq_len(length(nodes(rh_white))),
                            data = integer(length(nodes(rh_white))))
}
```

The helper `vol_to_surf()` returns a `NeuroSurface` with per-vertex labels
for the mapping geometry. We extract those labels and reuse them on the
inflated plotting geometry by simple concatenation.

```{r extract-labels}
lh_parcels <- lh_labels@data
rh_parcels <- rh_labels@data

# Combine into a single vector for add_surface_layer()
parcel_labels <- c(lh_parcels, rh_parcels)
```

## Build a multi-view surface plot

We now construct a `neurosurf_plot` with a bilaterally symmetric layout and
two views (`lateral` and `medial`). We first add a continuous map (here, the
network index derived from the parcel id) and then overlay the parcel outlines.

```{r build-plot}
# Derive a simple continuous map: network index from parcel id
network_index <- floor((parcel_labels - 1L) / (200 / 7)) + 1L
network_index[parcel_labels == 0] <- NA_integer_

p <- surface_plot(
  lh = lh,
  rh = rh,
  views  = c("lateral", "medial"),
  layout = "grid",
  zoom   = 2.5
)

# Add filled network layer with viridis-like map and a titled colourbar
p <- add_surface_layer(
  p,
  data         = network_index,
  cmap         = "viridis",
  color_range  = range(network_index, na.rm = TRUE),
  show_colorbar = TRUE,
  label        = "Network index",
  alpha        = 0.9
)

# Add parcel outlines with tasteful defaults
p <- add_atlas_outline(
  p,
  labels = parcel_labels,
  label  = "Schaefer-200",
  outline_offset = 0.5,
  outline_lwd    = 1.2
)
```

## Render a publication-style figure

Finally, we draw the figure with a horizontal colourbar beneath the grid and
slightly compressed colourbar height and spacing for a more compact layout.

```{r render-figure, fig.width=8, fig.height=4}
g <- draw_surface_plot(
  p,
  colorbar      = TRUE,
  cbar_location = "bottom",
  cbar_kws      = list(
    n_ticks    = 3,
    digits     = 0,
    label_cex  = 0.7,
    title_cex  = 0.9,
    bar_height = grid::unit(0.6, "lines"),
    bar_spacing = grid::unit(0.4, "lines")
  )
)

grid::grid.newpage()
grid::grid.draw(g)
```

The resulting figure shows:

- left and right inflated fsaverage surfaces;
- lateral and medial views;
- a continuous network index map with a clean colourbar; and
- crisp parcel outlines with a subtle halo and offset for legibility.

This pattern can be adapted to any surface + atlas combination that can be
loaded into neurosurf as `SurfaceGeometry` and per-vertex ROI labels.
