---
title: Publication-quality surface figures
author: neurosurf authors
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Publication-quality surface figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette builds a multi-view, publication-ready surface figure from
fsaverage surfaces and an atlas, using neurosurf's high-level plotting API:
`surface_plot()`, `add_surface_layer()`, `add_atlas_outline()`, and
`draw_surface_plot()`.

We will:

- load the bundled fsaverage `std.8` surfaces;
- map a Schaefer 200-parcel atlas from MNI152 volume space onto the surface;
- overlay a continuous statistical map with a colourbar; and
- add crisp atlas outlines for anatomical reference.

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (
  requireNamespace("systemfonts", quietly = TRUE) &&
  requireNamespace("albersdown", quietly = TRUE) &&
  "albers_register_fonts" %in% getNamespaceExports("albersdown")
) {
  getExportedValue("albersdown", "albers_register_fonts")()
}
# Prefer rgl's NULL device for non-interactive builds (avoids headless crashes)
if (!interactive()) {
  options(rgl.useNULL = TRUE)
}

# Skip the vignette gracefully if rgl is unavailable
has_rgl <- requireNamespace("rgl", quietly = TRUE)
if (!has_rgl) {
  knitr::opts_chunk$set(eval = FALSE)
  message("Skipping vignette: rgl package not available")
}

if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dev = "png",
  dev.args = list(bg = "white"),
  fig.width = 8,
  fig.height = 4.5,
  out.width = "100%"
)

library(neurosurf)
library(neuroim2)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


## Load surfaces and an atlas

The package ships decimated fsaverage surfaces (`std.8`). We use the
**inflated** surfaces for display, and the **white**/**pial** surfaces to
project the atlas from the volume onto the surface.

```{r load-surfaces}
fs_infl <- load_fsaverage_std8("inflated")
lh <- fs_infl$lh
rh <- fs_infl$rh

read_geom <- function(name) {
  read_surf_geometry(system.file("extdata", name, package = "neurosurf"))
}
lh_white <- read_geom("std.8_lh.white.asc")
lh_pial  <- read_geom("std.8_lh.pial.asc")
rh_white <- read_geom("std.8_rh.white.asc")
rh_pial  <- read_geom("std.8_rh.pial.asc")
```

`vol_to_surf()` maps each surface vertex to the most common atlas label among
the voxels between the white and pial surfaces, giving a per-vertex parcel id
for each hemisphere.

```{r map-atlas}
atlas <- neuroim2::read_vol(system.file(
  "extdata",
  "Schaefer2018_200Parcels_7Networks_order_FSLMNI152_1mm.nii.gz",
  package = "neurosurf"
))

lh_parcels <- as.integer(vol_to_surf(lh_white, lh_pial, atlas, fun = "mode")@data)
rh_parcels <- as.integer(vol_to_surf(rh_white, rh_pial, atlas, fun = "mode")@data)

# A single left-to-right vector matching the plotting geometry
parcel_labels <- c(lh_parcels, rh_parcels)
```

> **A note on resolution.** `std.8` is a heavily decimated mesh
> (`r length(nodes(lh))` vertices per hemisphere), so a 200-parcel atlas is
> represented coarsely. The outlines below trace whatever parcels land on the
> mesh; on a full-resolution surface the same code produces finer borders.

## A continuous overlay with a colourbar

We build a `neurosurf_plot` with a bilateral grid layout and two views
(`lateral` and `medial`), then add a continuous map. Here the map is a smooth
synthetic statistic standing in for, e.g., a contrast or connectivity value;
in practice you would pass your own per-vertex values.

```{r continuous-map}
# Smooth synthetic per-vertex statistic (replace with your own map)
example_statistic <- function(geom) {
  xyz <- coords(geom)
  v <- sin(xyz[, 1] / 18) + cos(xyz[, 2] / 22)
  (v - mean(v)) / stats::sd(v)
}
stat <- c(example_statistic(lh), example_statistic(rh))

p <- surface_plot(lh = lh, rh = rh,
                  views = c("lateral", "medial"),
                  layout = "grid", zoom = 2.5)

p <- add_surface_layer(
  p,
  data          = stat,
  cmap          = "viridis",
  show_colorbar = TRUE,
  label         = "Example statistic",
  alpha         = 0.95
)

plot(p, colorbar = TRUE, cbar_location = "bottom",
     cbar_kws = list(n_ticks = 3, digits = 1, label_cex = 0.7, title_cex = 0.9))
```

## Adding atlas outlines

`add_atlas_outline()` overlays parcel boundaries. By default it uses the
`"midpoint"` boundary method, which draws a single crisp contour running
between differing labels (no double lines), with a subtle halo for legibility
against the colour map.

```{r outlined-figure}
p <- add_atlas_outline(
  p,
  labels         = parcel_labels,
  label          = "Schaefer-200",
  outline_lwd    = 1.0,
  outline_offset = 0.5
)

plot(p, colorbar = TRUE, cbar_location = "bottom",
     cbar_kws = list(n_ticks = 3, digits = 1, label_cex = 0.7, title_cex = 0.9))
```

The figure now shows left and right inflated surfaces in lateral and medial
views, a continuous map with a clean colourbar, and parcel outlines for
anatomical reference. The same pattern works for any surface + atlas
combination that can be loaded as a `SurfaceGeometry` plus per-vertex labels.

## Next Steps

- `vignette("displaying-surfaces")` — lower-level RGL rendering with curvature shading, data overlays, thresholds, and PNG snapshots.
- `vignette("interactive-surfaces")` — interactive HTML widgets with `surfwidget()` for exploratory analysis.
- `vignette("introduction-to-neurosurf")` — the data structures (`SurfaceGeometry`, `NeuroSurface`, `NeuroSurfaceVector`) that underpin these workflows.
