---
title: Publication-quality continuous surface figures
author: neurosurf authors
date: '`r Sys.Date()`'
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: teal
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Publication-quality continuous surface figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) {
  knitr::opts_chunk$set(dev = "ragg_png")
}
if (
  requireNamespace("systemfonts", quietly = TRUE) &&
  requireNamespace("albersdown", quietly = TRUE) &&
  "albers_register_fonts" %in% getNamespaceExports("albersdown")
) {
  getExportedValue("albersdown", "albers_register_fonts")()
}
if (requireNamespace("ggplot2", quietly = TRUE) &&
    requireNamespace("albersdown", quietly = TRUE)) {
  ggplot2::theme_set(
    albersdown::theme_albers(
      family = params$family,
      preset = params$preset
    )
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)

library(neurosurf)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

You have a continuous value at every cortical vertex--for example a *t*, *z*,
thickness, or connectivity map--and need a static figure that remains
scientifically interpretable after thresholding. The important distinction is
that the value is a **field on a triangle mesh**, not a parcel label. A polished
figure should therefore interpolate the scalar within each triangle, test
visibility with a real depth buffer, and add anatomy without drawing atlas or
occlusion lines over the result.

`render_surface_rgba()` is neurosurf's deterministic, headless primitive for
that task. It returns an RGBA raster plus diagnostic buffers and provenance;
`write_surface_rgba()` writes the result to PNG. Higher-level packages can
compose several such panels with vector labels and legends.

## What are the inputs?

You need a display geometry and one numeric value per vertex. For anatomical
context, use a curvature or sulcal-depth metric from a corresponding white,
pial, or midthickness surface rather than computing curvature on the inflated
display shape.

The package includes matching decimated fsaverage surfaces, which keep this
example fast and offline.

```{r load-surfaces}
inflated <- load_fsaverage_std8("inflated")

read_geom <- function(name) {
  read_surf_geometry(system.file("extdata", name, package = "neurosurf"))
}
white_lh <- read_geom("std.8_lh.white.asc")
white_rh <- read_geom("std.8_rh.white.asc")
```

Here a smooth coordinate-based field stands in for your measured statistic.
The renderer treats it exactly like any other vertex-wise numeric vector.

```{r make-statistic}
example_statistic <- function(geometry) {
  xyz <- coords(geometry)
  value <- sin(xyz[, 2] / 18) + cos(xyz[, 3] / 14) +
    0.7 * sin(xyz[, 1] / 12)
  as.numeric(scale(value))
}

stat_lh <- example_statistic(inflated$lh)
stat_rh <- example_statistic(inflated$rh)
```

```{r validate-inputs, include=FALSE}
curv_lh <- curvature(white_lh)
curv_rh <- curvature(white_rh)
stopifnot(
  length(stat_lh) == length(curv_lh),
  length(stat_rh) == length(curv_rh),
  all(is.finite(stat_lh)),
  all(is.finite(stat_rh)),
  stats::sd(stat_lh) > 0,
  stats::sd(stat_rh) > 0
)
```

## How do you render one thresholded view?

Choose the threshold and colour limits as scientific inputs. In this example,
the colour scale stays fixed at `[-2.5, 2.5]`, while an opacity ramp softens
only the visual transition immediately above `|z| = 1`.

```{r render-one-view}
lh_lateral <- render_surface_rgba(
  inflated$lh,
  vertex_values = stat_lh,
  anatomy_metric = curv_lh,
  camera = "lateral",
  threshold = 1,
  tail = "two_sided",
  limits = c(-2.5, 2.5),
  alpha_ramp = 0.25,
  antialias = 2
)
```

```{r validate-one-view, include=FALSE}
active <- as.integer(lh_lateral$overlay_alpha) > 0
stopifnot(
  any(lh_lateral$coverage),
  any(active),
  any(lh_lateral$coverage & !active),
  identical(lh_lateral$provenance$backend, "cpu_barycentric"),
  identical(lh_lateral$provenance$limits, c(-2.5, 2.5))
)
```

```{r show-one-view, echo=FALSE, out.width="80%"}
one_view_file <- knitr::fig_path("-cpu-lh-lateral.png")
dir.create(dirname(one_view_file), recursive = TRUE, showWarnings = FALSE)
write_surface_rgba(lh_lateral, one_view_file)
knitr::include_graphics(one_view_file)
```

The colour at each covered sample comes from the barycentrically interpolated
scalar. Thresholding and palette mapping happen after that interpolation, so a
threshold boundary may cross a triangle instead of jumping from one mesh edge
to the next. The per-sample z-buffer independently decides which folded surface
fragment is visible.

## How do you build the four canonical views?

Use the same limits, threshold, opacity, and anatomy convention in every
panel. The camera contract is explicit: `camera_mode = "canonical"` is strict
orthographic output, while `"presentation"` adds a small declared obliquity.

```{r render-four-views, include=FALSE}
render_panel <- function(geometry, statistic, anatomy, view) {
  render_surface_rgba(
    geometry,
    vertex_values = statistic,
    anatomy_metric = anatomy,
    camera = view,
    camera_mode = "canonical",
    threshold = 1,
    tail = "two_sided",
    limits = c(-2.5, 2.5),
    alpha_ramp = 0.25,
    width = 720,
    height = 450,
    antialias = 2
  )
}

panels <- list(
  lh_lateral = render_panel(inflated$lh, stat_lh, curv_lh, "lateral"),
  rh_lateral = render_panel(inflated$rh, stat_rh, curv_rh, "lateral"),
  lh_medial = render_panel(inflated$lh, stat_lh, curv_lh, "medial"),
  rh_medial = render_panel(inflated$rh, stat_rh, curv_rh, "medial")
)

panel_files <- vapply(names(panels), function(name) {
  path <- knitr::fig_path(paste0("-cpu-", name, ".png"))
  dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
  write_surface_rgba(panels[[name]], path)
  path
}, character(1))

stopifnot(
  all(vapply(panels, function(x) any(x$coverage), logical(1))),
  all(vapply(panels, function(x) any(as.integer(x$overlay_alpha) > 0),
             logical(1)))
)
```

```{r show-four-views, echo=FALSE, out.width="49%", fig.show="hold"}
knitr::include_graphics(panel_files)
```

The lateral and medial views differ because each panel rasterizes the visible
triangles from its own camera. Internal sulcal occlusion edges are not drawn as
outlines; `outer_contour = TRUE` marks only cortex touching background that is
connected to the image exterior.

## Where do the medial wall and cortex mask enter?

A real cortex mask is a separate domain object, not a consequence of parcel
value zero. Pass one logical value per vertex. Triangles touching a masked
vertex never receive overlay colour. `medial_wall` then controls whether the
excluded domain is quietly shaded, omitted, or outlined.

```{r cortex-mask-contract, eval=FALSE}
masked_panel <- render_surface_rgba(
  inflated$lh,
  vertex_values = statistic,
  anatomy_metric = white_surface_curvature,
  cortex_mask = cortex_label,
  medial_wall = "shade"
)
```

Atlas-aware callers such as `neuroatlas::plot_brain()` can resolve the mask and
its provenance from an atlas annotation. At this low level, neurosurf requires
the caller to supply the domain explicitly rather than guessing that an atlas
label of zero always means medial wall.

## How should a volume become a continuous surface field?

For a continuous image, use explicit trilinear sampling through the cortical
ribbon. Interpolation, aggregation across depth, and tangential smoothing are
separate choices. The following contract samples five depths, averages them,
and applies no surface smoothing:

```{r ribbon-projection, eval=FALSE}
mapped <- vol_to_surf(
  white_surface,
  pial_surface,
  statistic_volume,
  fun = "avg",
  sampling = "thickness",
  interpolation = "linear",
  depth = seq(0.1, 0.9, length.out = 5),
  aggregate = "mean",
  surface_smooth_fwhm = 0
)
```

Use `interpolation = "nearest"` for discrete nearest-voxel sampling. The
historical Gaussian KNN behaviour remains available as
`interpolation = "legacy"`. Categorical `aggregate = "mode"` is intentionally
incompatible with linear interpolation.

## Which renderer should you use?

- Use `render_surface_rgba()` for deterministic, scalar-first static output in
  CI, Slurm, Quarto, or PDF workflows. It requires neither OpenGL nor a
  browser.
- Use `surface_plot()`, `view_surface()`, or `surfwidget()` for interactive
  exploration, atlas inspection, and report widgets. Those paths provide rich
  3D interaction, but they are not the scalar-threshold oracle for static
  continuous maps.
- Use `neuroatlas::plot_brain(style = "stat_publication",
  static_backend = "cpu")` when you need anatomical masks, curvature
  resolution, a four-view layout, orientation marks, and an overlay legend.
- Use `neuromosaic::surf_montage()` when the starting point is a statistic
  volume and the figure belongs in a reproducible report.

Atlas outlines remain useful when the atlas itself is the subject of the
figure. They are deliberately not part of the default continuous-statistic
view, where parcel boundaries would imply structure that is absent from the
scalar field.

## Next steps

- `vignette("displaying-surfaces")` covers lower-level RGL rendering and local
  snapshots.
- `vignette("interactive-surfaces")` builds interactive bilateral report
  widgets with `surfwidget()`.
- `vignette("introduction-to-neurosurf")` introduces `SurfaceGeometry`,
  `NeuroSurface`, and related data structures.
- `?render_surface_rgba`, `?surface_threshold_segments`, and `?vol_to_surf`
  document the complete computational contracts.
